Thesis Proposal

### 2019-09-19 ThesisProposal.Rmd
Update after feedback Gerko on draft thesis proposal

Supervised by Prof. Dr. Stef van Buuren, & Dr. Gerko Vink

@@

# Introduction/context and research question/goals

At some point, any (social) scientist conducting statistical analyses will run into a missing data problem (@alli02). Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences (@buur18). Multiple imputation (MI; @rubin87) provides a framework to circumvent the problem of missing information. This technique -- that is growing in popularity (@buur18) -- entails completing the missing values in an incomplete data set several times. The variability between the completed data sets captures the uncertainty in the inferences due to missingness (@rubin87).

The R package MICE (Multiple Imputation using Chained Equations; @mice11) is the world-leading software for multiple imputation (@buur18). However, the package currently does not provide user-friendly means to evaluate the multiply imputed data. Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence). The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness. Simultaniously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms.

@@

Currently, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5). Instead, users have to rely on visual inspection of the algorithm's iterations. Convergence is said to be reached when the parameters (like means of completed variables or regression coefficients) are stable over iterations (@whit11). 
<!-- "As MICE is an iterative procedure, it is important that convergence is achieved. This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p. 394). -->
Additionally, convergence requires the variance between the several imputation chains to be no larger than the variances within each chain (@buur18). 
<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
According to @li14 the latter is conceptually similar to the Gelman-Rubin statistic. 
<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3). -->

The Gelman-Rubin statistic (@gelm92), however, assumes independence between chains. Therefore, it cannot be directly applied to compare imputations that share the same incomplete data (source??).

Notwithstandlingly, it seems like @su11 did implement the Gelman-Rubin statistic as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure. By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation. If the $\hat{R}$ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004). Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p. 13). -->
It is worth investigating its validity in the context of MICE's convergence, and the broader framework of fully conditional specification (FSC) algorithms.

@@

@@

During this research project I will work directly with developers of the MICE R package. I will use R Shiny (@shiny17) to program a web-browser based local app for the evaluation of multiply imputed data: 'ShinyMICE'. This approach does not require the use of any (personal) data. Therefore, the project will not be reviewed by the ethical committee.

@@

Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vingette for applied researchers will be written.

@@

"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11). It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p. 19).

>>>>

Supervised by prof. dr. Stef van Buuren, & dr. Gerko Vink

@@

# Introduction

At some point, any (social) scientist conducting statistical analyses will run into a missing data problem (@alli02). Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences (@buur18). Multiple imputation (MI; @rubin87) provides a framework to circumvent the *ubiquitous* problem of missing information. This technique -- that is growing in popularity (@buur18) -- entails 'guessing' the missing values in an incomplete data set several times. The variability between the resulting completed data sets represents how much uncertainty in the inferences is due to missingness (@rubin87).

The R package MICE (Multiple Imputation using Chained Equations; @mice11) is the world-leading software for multiple imputation (@buur18). However, the package currently does not provide user-friendly means to evaluate multiply imputed data. *This is vital because of the amount of assumptions in MI algorithms. Thanks to technological developments in computing speed and data storage it has become possible to check assumptions where that was unfeasible before. The very first assumption to check is algorithmic convergence; without convergence, all 'deeper' assumptions and subsequent inferences are invalid.*

Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence). The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness. Simultaneously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms. *The research question central to this project is: 'How can empirical researchers be facilitated in evaluating the  validity of multiply imputed data?'.*

@@

Currently, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5). Instead, users have to rely on visual inspection of the algorithm's iterations. Convergence is said to be reached when parameters (like means of completed variables or regression coefficients) are stable across iterations (@whit11). 
<!-- "As MICE is an iterative procedure, it is important that convergence is achieved. This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p. 394). -->
And additionally, the variation between imputations should be no larger than the variation across iterations of each individual chain (@buur18). 
<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
*Conceptually, this visual evaluation procedure is similar to the 'Gelman-Rubin statistic' (@li14). Practically, however, Gelman and Rubin's convergence diagnostic $\widehat{R}$ ('potential scale reduction'; 1992) cannot be directly applied to assess convergence of MICE algorithms because the measure is famously prone to produce false positives when applied to MI data.*  
<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3). -->
*That is, the overdispersed initial state of the MICE algorithm does not conform to $\widehat{R}$'s assumption of equal starting points (source!).*

Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure. By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation. If the $\hat{R}$ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004). Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p. 13). -->
It is worth investigating the validity of the convergence statistic in the context of MICE, and the broader framework of fully conditional specification (FSC) algorithms.

@@

**Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see @veht19. Or leave it for the Research Report.**

@@

During this research project I will work directly with developers of the MICE R package. I will use R Shiny (@shiny17) to program an interactive evaluation device for the evaluation of multiply imputed data: 'ShinyMICE'. This approach does not require the use of any empirical data -- personal or otherwise. Therefore, the project does not need to be reviewed/approved by the faculty's ethical committee.

@@

Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.

@@

"For models based on multiple imputed data sets, this [R hat > 1.1] is often a false positive: Chains of different submodels may not overlay each other exactly, since there were fitted to different data" (see https://cran.r-project.org/web/packages/brms/vignettes/brms_missings.html).

"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11). It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p. 19).
